---
title: "Class 7: Machine Learning I"
author: "Youn Soo Na (PID: A1704731)"
format: pdf
---

> Today we are going to learn how to apply different machine learning methods, beginning with clustering:

> The goal here is to find groups/clusters in your input data.

> First, I will make up some data with clear groups. For this I will use the `rnorm()` function:

```{r}
rnorm(10)
```

```{r}
hist( rnorm(10000, mean = 3) )
```

```{r}
n <- 10000
x <- c(rnorm(n,-3), rnorm(n, +3))
hist(x)
```

```{r}
n <- 30
x <- c(rnorm(n,-3), rnorm(n, +3))
y <- rev(x)
z <- cbind(x,y)
head(z)
plot(z)
```


> Use the `kmeans()` function setting k to 2 and nstart=20

Inspect/print the results

> Q. How many points are in each cluster?

> Q. What 'component' of your result object details
        - cluster size?
        - cluster assignment/membership?
        - cluster center?
        


```{r}
km <- kmeans(z, centers = 2)
km
```

Results in kemans object `km`
```{r}
attributes(km)
```

cluster size?

```{r}
km$size
```

cluster assignment/membership?
```{r}
km$cluster
```

cluster center?
```{r}
km$centers
```

> Q. Plot x colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
plot(z, col = c("red", "blue"))
```

R will re-cycle the shorter color vector to be the same length as the longer (number of data poitns) in z. 

```{r}
plot(z, col = km$cluster)
```

We can use the `points()` function to add new points to an existing plot. . .

```{r}
plot(z, col = km$cluster)
points(km$centers, col="blue", pch=33, cex = 3)
# max pch is 127
```

> Q. Can you run kmeans and ask for 4 clusters please and plot the results like we have done above?

```{r}
km4 <- kmeans(z, centers = 4)
plot(z, col = km4$cluster)
points(km4$centers, col="blue", pch=33, cex = 3)
```


## Hierarchical Clustering

Let's take our same made-up data `z` and see how hclust works.

First we need a distance matrix of our data to be clustered.

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

```{r}
plot(hc)
```

I can get my cluster membership vector by "cutting the tree" with the `cutree()` function like so

```{r}
grps <- cutree(hc, h=8)
grps
```

Can you plot `z` colored by out hclust results?

```{r}
plot(z, col=grps)
```

## PCA of UK Food Data

Read data from the UK on food consumption in different parts of the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
head(x)
```
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

A so-called "Pairs" plot can be useful for small datasets like this

```{r}
pairs(x, col=rainbow(10), pch=16)
```

It's hard to see structure and trends in even this small data-set. How will we ever do this when we have big data-sets with 1,000s or 10s of thousands of things we are measuring. . .


## PCA to the rescue

Let's see how PCA deals with this dataset. So main function in base R to do PCA is called `prcomp()`

```{r}
#transpose t()
pca <- prcomp( t(x) )
summary(pca)
```

Let's see what is inside this `pca` object that we created from running `prcomp()`

```{r}
attributes(pca)
```

```{r}
pca$x
```

```{r}
plot(pca$x[,1], pca$x[,2], col = c("black", "red", "blue", "darkgreen"), pch=16,
     xlab="PC1 (67.4%)", ylab = "PC2 (29.0%)")
text(pca$x[,1], pca$x[,2], labels=c("England", "Wales", "Scotland", "N.Ireland"), col=c("black", "red", "blue", "darkgreen"))

```

## Variable Loadings Plot

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```







