---
title: "Lab 12"
author: "Na Youn Soo (PID: A17014731)"
format: gfm
---


> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
file <- read.table("rs8067378_ENSG00000172057.6.txt")
```

```{r}
summary(file)
# Sample size for each genotype
table(file$geno)
```

```{r}
aa <- subset(file, geno == "A/A")
# Median Expression Levels of genotype A/A samples
median(aa$exp)
```
```{r}
ag <- subset(file, geno == "A/G")
# Median Expression Levels of genotype A/G samples
median(ag$exp)
```
```{r}
gg <- subset(file, geno == "G/G")
# Median Expression Levels of genotype G/G samples
median(gg$exp)
```


> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(file, aes(geno, exp, fill = geno)) +
         geom_boxplot() +
  guides(fill="none") +
    geom_jitter(width = 0.2, alpha = 0.2) +
  labs(title="Per Genotype Expression Values", x = "Genotype", y = "Expression")
```

A/A is expressed more than G/G

Given the large difference in expression between the genotypes, I believe SNP affects the expression of ORMDL3.










